The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-07-01 paper/iphas-dr2-rev1.tex
Initial revisions

When the initial data release was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to 92 per cent of the survey area 
and includes a global photometric re-calibration.
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we recap some key points.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r$ and $i$ 
is extracted along with narrowband \ha. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

@@

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and the source code is available
on-line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

@@

At the time of preparing DR2,
the CASU pipeline had processed
74,195 single-band IPHAS exposures 
in which a total of 1.9~billion \emph{candidate} detections
were made at the default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious objects, artefacts and
duplicate detections;
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a more reliable source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

\begin{figure}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/caldiagram/ccd-uncalibrated.pdf}
\caption{}
\label{fig:ccd_before}
\end{subfigure}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/caldiagram/ccd-calibrated.pdf}
\caption{}
\label{fig:ccd_after}
\end{subfigure}
\caption{IPHAS $(r-\ha,\ r-i)$ colour-colour diagram
         covering an area of 400~deg$^2$,
         shown before (panel a) and after (panel b)
         re-calibration.
         Both figures were created by combining the stars
         detected across all 2,801 quality-approved fields
         which are located towards the Galactic anti-centre
         ($160\degr<\ell<200\degr$).
         The diagrams are plotted as 2D-histograms
         which show the density of sources
         in bins of 0.01-by-0.01~mag;
         bins containing 1 to 200 sources are coloured red,
         while bins containing more than 400 sources are bright yellow.
         The diagrams include all stars
         brighter than $r<18$ which were
         classified by the pipeline as being `very reliable'
         (to be explained in
         \S\ref{sec:qualitycriteria}).
         The objects which are seen to fall above the locus of stars
         \emph{after} re-calibration 
         are likely to be genuine \ha\ emission-line objects.}
\end{figure}

@@

Having re-calibrated all fields to the expected APASS accuracy of 3 per cent,
we turned to a different survey, SDSS Data Release 9 \citep{Ahn2012},
to validate the results.
SDSS DR9 includes several strips at low
Galactic latitudes,
providing data across 18 per cent of the fields in our data release.
We cross-matched the IPHAS fields against the subset of
objects marked as reliable stars in the SDSS catalogue\footnote{
We used the CasJobs facility located at http://skyserver.sdss3.org/CasJobs
to obtain photometry from the SDSS {\sc photoprimary} table 
with criteria {\sc type = star}, {\sc clean = 1} and {\sc score $>$ 0.7}.}
in much the same way as for APASS,
with the difference of selecting from the  fainter magnitude ranges of 
$15<r_{\rm SDSS}<18.0$ and $14.5<i_{\rm SDSS}<17.5$.
This provided us with a set of 1.2~million cross-matched stars.

@@

Table~\ref{tab:reliable} specifies
a set of minimum quality criteria
which should benefit most applications
which desire reliable colours
as well as completeness.
The listed criteria are designed to 
(i) remove low-S/N sources, 
(ii) remove saturated sources,
and (iii) remove objects for which the 2\farcs 3-diameter
aperture magnitude is inconsistent 
with its alternative 1\farcs 2-diameter measurement
within each filter.
This last criterion is a proxy
for identifying objects which are affected
by poor background subtraction
or failed source deblending.
A total of 86 million sources out of 219 million 
(39 per cent) satisfy all the criteria listed in Table~\ref{tab:reliable}
and are hereafter referred to as `reliable'.
For convenience, the catalogue contains a boolean column
named \emph{reliable} that directly flags these objects.

@@

\begin{figure}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{figures/repeatability/repeatability.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{figures/repeatability/repeatability-reliable.pdf}
\caption{}
\label{fig:pairmag_reliable}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for all sources in the catalogue (left panel)
         and for the \emph{veryReliable} sources alone (right panel).
         Black dots show the mean absolute residuals
         between the primary and the secondary detections.
         The error-bars show the standard deviations.
         The red trend line shows the average Poissonian uncertainties
         we derived in Fig.~\ref{fig:uncertainties}.
         We find that the \emph{veryReliable} quality criteria are successful
         at removing objects with large residuals.
    }
\end{figure}
\begin{figure*}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{figures/sourcecount/sourcecount.pdf} 
    \caption{Number of entries in the IPHAS DR2 source catalogue
    as a function of Galactic longitude.
    The upper blue line shows the number of sources
    counted in 1\degr-wide longitude bins.
    The lower red line uses the same binning
    but includes only those sources 
    for which the \emph{deblend} flag is {\sc false}, 
    i.e. unconfused sources for which the CASU pipeline
    did not have to apply a deblending procedure.
    In both cases we counted only those sources
    in the latitude range $-5\degr<b<+5\degr$,
    such that one may obtain a rough guide to source density
    by dividing the counts by 10~deg$^2$.
    The global variations in the source counts
    traces the structure of the Galaxy
    and the distribution of foreground extinction,
    but is also affected by instrumental effects
	such as variations in the observed depth
	and completeness (see text).
   }
    \label{fig:sourcecount}
\end{figure*}

>>>>

{\bf When the previous release -- the IDR -- was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release, 
 which supersedes the IDR by providing a global re-calibration
and taking the coverage up to 92 per cent of the survey area.}
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

\begin{table}
    \caption{Key properties of IPHAS DR2.}
    \label{tbl:survey}
    \begin{center}
        \begin{tabular}{ll}
        \toprule
        Property & Value \\
        \midrule
        Telescope & 2.5m Isaac Newton Telescope, La Palma \\
        Instrument & Wide Field Camera \\
        Detectors & Four 2048 $\times$ 4100 pixel CCD's \\
        %Detector field of view & 11 $\times$ 23 arcmin \\
        Detector scale & 0.33 arcsec/pixel \\        
        Filters & $r$, $i$, \ha \\
        Filter properties & See Table~2 \\
        Exposure times & 30~s ($r$), 10~s ($i$), 120~s (\ha) \\
		Saturation limit & TBC \\
        Detection limit (5$\sigma$, Vega) & 21.3 ($r$), 20.1 ($i$), 20.3 (\ha) \\
        PSF FWHM (median) & 1.1\arcsec\ ($r$), 1.0\arcsec\ ($i$), 1.1\arcsec\ (\ha) \\
        Survey area & $\sim 1860$~deg$^2$ \\
        Footprint boundaries & $ -5\degr < b < +5\degr,\ 29\degr < \ell < 215\degr$ \\
        DR2 observing period & August 2003 - November 2012 \\
        Website & www.iphas.org \\
        \bottomrule
        \end{tabular}
	\end{center}
\end{table}

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release,
we recap some key points {\bf below and in Table~1}.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r$ and $i$ 
is extracted along with narrowband \ha. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

@@

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archiving.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and the source code is available
on-line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

@@

At the time of preparing DR2,
the CASU pipeline had processed
74,195 single-band IPHAS exposures 
in which a total of 1.9~billion \emph{candidate} detections
were made at the default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious objects, artefacts and
duplicate detections;
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a more reliable source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
{ \bf It is these post-processing steps which distinguish the present release
from the IDR, which enforced less stringent quality limits,
did not offer a global calibration,
and provided a catalogue in which duplicate detections
of unique sources were not flagged.
These new processing steps are explained next.}
%It is these post-processing steps which are explained next.

@@

\begin{figure}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/caldiagram/ccd-uncalibrated.pdf}
\caption{}
\label{fig:ccd_before}
\end{subfigure}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/caldiagram/ccd-calibrated.pdf}
\caption{}
\label{fig:ccd_after}
\end{subfigure}
\caption{IPHAS $(r-\ha,\ r-i)$ colour-colour diagram
         covering an area of 400~deg$^2$,
         shown before (panel a) and after (panel b)
         re-calibration.
         Both figures were created by combining the stars
         detected across all 2,801 quality-approved fields
         which are located towards the Galactic anti-centre
         ($160\degr<\ell<200\degr$).
         The diagrams are plotted as 2D-histograms
         which show the density of sources
         in bins of 0.01-by-0.01~mag;
         bins containing 1 to 200 sources are coloured red,
         while bins containing more than 400 sources are bright yellow.
         {\bf The diagrams include all stars
         brighter than $r<18$ which were
         classified by the pipeline as `bestQualityAllBand'
         (indicating a high-significance point source detection with
         accurate photometry in all bands,
         to be explained in \S\ref{sec:qualitycriteria})}.
         The objects which are seen to fall above the locus of stars
         \emph{after} re-calibration 
         are likely to be genuine \ha\ emission-line objects.}
\end{figure}

@@

Having re-calibrated all fields to the expected APASS accuracy of 3 per cent,
we turned to a different survey, SDSS Data Release 9 \citep{Ahn2012},
to validate the results.
{\bf SDSS DR9 includes photometry across several strips at low
Galactic latitudes,
which were observed as part of 
the Sloan Extension for Galactic Understanding and Exploration \citep[SEGUE;][]{segue}.}
These strips provide data across 18 per cent of the fields in our data release.
We cross-matched the IPHAS fields against the subset of
objects marked as reliable stars in the SDSS catalogue\footnote{
We used the CasJobs facility located at http://skyserver.sdss3.org/CasJobs
to obtain photometry from the SDSS {\sc photoprimary} table 
with criteria {\sc type = star}, {\sc clean = 1} and {\sc score $>$ 0.7}.}
in much the same way as for APASS,
with the difference of selecting from the  fainter magnitude ranges of 
$15<r_{\rm SDSS}<18.0$ and $14.5<i_{\rm SDSS}<17.5$.
This provided us with a set of 1.2~million cross-matched stars.

@@

Table~\ref{tab:reliable} specifies
a set of minimum quality criteria
which should benefit most applications
which desire reliable colours
as well as completeness {\bf down to $\sim$19th magnitude}.
The listed criteria are designed to 
(i) remove low-S/N sources, 
(ii) remove saturated sources,
and (iii) remove objects for which the 2\farcs 3-diameter
aperture magnitude is inconsistent 
with its alternative 1\farcs 2-diameter measurement
within each filter.
This last criterion is a proxy
for identifying objects which are affected
by poor background subtraction
or failed source deblending.
A total of 86 million sources out of 219 million 
(39 per cent) satisfy all the criteria listed in Table~\ref{tab:reliable}
and are hereafter referred to as {\bf `highQualityAllBand'}.
For convenience, the catalogue contains a boolean column
named {\bf \emph{highQualityAllBand}} that directly flags these objects.

@@

\begin{figure}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{figures/repeatability/repeatability.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{figures/repeatability/repeatability-reliable.pdf}
\caption{}
\label{fig:pairmag_reliable}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for all sources in the catalogue (panel a)
         and for the \emph{veryReliable} sources alone (panel b).
         Black dots show the mean absolute residuals
         between the primary and the secondary detections.
         The error-bars show the standard deviations.
         The red trend line shows the average Poissonian uncertainties
         we derived in Fig.~\ref{fig:uncertainties}.
         We find that the \emph{veryReliable} quality criteria are successful
         at removing objects with large residuals.
    }
\end{figure}
\begin{figure*}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{figures/sourcecount/sourcecount.pdf} 
    \caption{Number of entries in the IPHAS DR2 source catalogue
    as a function of Galactic longitude.
    The upper blue line shows the number of sources
    counted in 1\degr-wide longitude bins.
    The lower red line uses the same binning
    but includes only those sources 
    for which the \emph{deblend} flag is {\sc false}, 
    i.e. unconfused sources for which the CASU pipeline
    did not have to apply a deblending procedure.
    In both cases we counted only those sources
    in the latitude range $-5\degr<b<+5\degr$,
    such that one may obtain a rough guide to source density
    by dividing the counts by 10~deg$^2$.
    The global variations in the source counts
    traces the structure of the Galaxy
    and the distribution of foreground extinction,
    but is also affected by instrumental effects
	such as variations in the observed depth
	and completeness (see text).
   }
    \label{fig:sourcecount}
\end{figure*}
